Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-01 Hyperplane Enumeration/main.tex
More hyperplane updates

A critical aspect of this problem class is that it can be optimized using Gray-Box
techniques, which can allow for highly efficient local~\cite{whitley:2013:greedy}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
exploit the known and limited non-linearity between variables expressed by all $s_i$,
as well as perform partial solution evaluation using all $f_i$.

@@

TODO Explain all problems here

@@

\section{Gray-Box Enumeration}
TODO Explain how gray-box properties and gray coding can be used to
enumerate in \BigO{2^N} time.~\cite{ochoa:2015:crossovernetworks}

@@

@@

>>>>

A critical aspect of this problem class is that it can be optimized using Gray-Box
techniques, which can allow for highly efficient local~\cite{whitley:2012:constant}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search. These algorithms
exploit the known and limited non-linearity between variables expressed by all $s_i$,
as well as perform partial solution evaluation using all $f_i$.

@@

NKq landscapes specify a class of randomly generated problem instances using 3 paramters:
(1) the number of problem variables $N$ (2) the amount of variable epistasis $K$ where $K=k-1$
and (3) the number of unique fitness values $q$. From these parameters a landscape is generated
by creating $M=N$ subfunctions $f_i$, where $f_i$ uses variable $i$ and $K$ others to look up
a fitness value in the range $[0..q-1]$ from a randomly generated table. This structure
makes NKq landscapes a natural fit for Mk landscapes (sum of bounded subfunctions)
and are the most studied problem class for Gray-Box
optimization~\cite{whitley:2012:constant,goldman:2015:GBO,tintos:2015:partitioncross,ochoa:2015:crossovernetworks,whitley:2015:mk}.
Here we choose $k=3$ with $q=2^{K+1}=2^{k}=8$.

We use two common variants of NKq in our experiments which specify how the $K$ additional
variables in each subfunction are chosen: Adjacent and Random. In Adjacent NKq, $f_i$ depends
on variables $[i..(i+k)\%N]$. In Random NKq, $f_i$ depends on variables $i$ and a random
set of $K$ unique variables which does not include $i$. While Random NKq is NP-Hard,
the structure of Adjacent NKq allows for a polynomial time dynamic programming solution~\cite{wright:2000:solvingnk}.
Adjacent NKq is therefore often used to understand a search algorithm's ability to reach the known global
optimum.

Ising Spin Glasses are a type of MAX-CUT problem derived from statistical physics.
Each atom in the glass (vertex) can be assigned a spin, with the goal being to
find the set of assignments which minimize the energy between nearby atoms (edges).
Similar to Adjacent NKq, the $2D\pm J$ subset of Ising Spin Glasses can be polynomially
solved~\cite{saul:1994:spinglass}\footnote{\url{http://www.informatik.uni-koeln.de/spinglass/}}.
In this subset, the graph is defined as a square two-dimensional torus with each edge
weight chosen from ${-1, 1}$. Each vertex is assigned a spin from ${-1, 1}$ with
the energy in the glass equal to
\begin{equation}
\sum_{e_{ij} \in E} x_ie_{ij}x_j
  \label{eq-ising}
\end{equation}
where $e_{ij}$ is the edge connecting vertex $i$ to vertex $j$. In Mk landscape terms
this type of spin glass has $M=2N$ and $k=2$.

Our final examined problem is randomly generated maximum satisfiability or MAX-kSAT.
This version of the canonical NP-Complete boolean satisfiability problem is formulated
as the maximization of $M=4.27N$ clauses, each containing exactly $k=3$ unique literals.
A clause is satisfied if any of its literals match how a solution's variables are set.

\section{Gray-Box Enumeration}
When considered as a black box, the process of finding all local optima
in a landscape requires \BigO{N2^N} time. This is because evaluating
a solution requires $N$ time, and all $2^N$ solutions in the landscape
must be evaluated once. Finding the local
optima then requires each solution to compare its fitness with each
of its neighbors, which when properly cached takes \BigO{N} per solution.
Extending this method to look for solutions which cannot be improved by
flipping $r$ or less bits requires \BigO{N^r2^N} time. However,
by exploiting Gray-Box optimization methods previous work~\cite{ochoa:2015:crossovernetworks}
was able to find all $r$-bit local optima of Mk Landscapes in \BigO{2^N} for any small,
fixed value of $r$.

The first major result in Gray-Box optimization was the proof that the list
of fitness improving moves during local search can be kept updated in \BigO{1}
time~\cite{whitley:2012:constant}. Each time a modification is made to a solution
only modifications which share a non-linear relationship with the changes can
go from non-improving to improving and vice versa. In Gray-Box optimization
the non-linearity between variables is known. Furthermore the number of
non-linear relationships is bounded above by \BigO{Mk} as each of the $M$
subfunctions can cause a non-linear relationship between each of its $k$
variables. As $M=$\BigO{N} and $k$ is constant, the amortized number of affected
flips per variable is \BigO{1}.

This limited and known non-linearity can be used to allow search to find only $r$-bit
local optima. Consider a variable interaction graph, such that each variable maps
to a vertex in the graph and each edge represents a non-linear relationship between
variables. TODO

@@

@@

\subsection{OneMax}
In OneMax, each bin contains exactly 1 move. All variables in the current solution
which disagree with the global optimum will be fitness improving moves.
Initially $index$ is set to $N-1$ and is decreased by at most 1 in any step.
Before finding the global optimum no carry operations can occur, meaning the
cost of each iteration is equal to the amount by which $index$ is decreased.
Therefore, Algorithm~\ref{alg-enumerate} requires \BigO{N} time to reach
the global optimum.

After finding the global optimum of OneMax, $index$ is set to 0. One iteration
is then spent adding a 1 to $index$ 0, which in the worst case requires \BigO{N}
carry operations. For all future iterations $index$ is a fitness improving move
and is currently set to 1. In these iterations at least 1 carry operation must
occur, and $index$ cannot be decreased. Iteration ends when $index$ exceeds $N$
which requires at most $N$ carry operations. Therefore, Algorithm~\ref{alg-enumerate}
requires \BigO{N} time to reach termination after finding the global optimum. Combined
with initialization and the time to find the global optimum, the total complexity is \BigO{N}.

\subsection{Concatenated Traps}
TODO

@@

TODO Consider discussing neutral networks
